---
title: "Reproducible Research Project 1"
author: "kev716"
date: "September 14, 2017"
output: html_document
---

# Reproducible Research Project 1

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## 1. Code for reading in the dataset and/or processing the data

```{r}

# check if activty.csv is in workspace, if not unzip project file
if (!file.exists("activity.csv")) {
  unzip("repdata%2Fdata%2Factivity.zip")
}

# read actity.csv into dataframe
  stepsDf <- read.csv("activity.csv")

# convert date field to date
stepsDf$date <- as.Date(stepsDf$date)
```

## 2. Histogram of the total number of steps taken each day

```{r}
# calculate the number steps per day
stepsDaily <- tapply(stepsDf$steps, stepsDf$date, sum, na.rm=TRUE)

# change resulting array into dataframe
stepsDaily <- as.data.frame(stepsDaily)

# change column name
names(stepsDaily) <- "steps"

# plot histogram of number of steps per day
qplot(stepsDaily$steps, geom="histogram")
```

## 3. Mean and median number of steps taken each day

```{r}

meanSteps <- mean(stepsDaily$steps)
meanSteps

medSteps <- median(stepsDaily$steps)
medSteps
```

## 4. Time series plot of the average number of steps taken per interval

```{r}

# calculate mean number of steps per interval
stepsInt <- tapply(stepsDf$steps, stepsDf$interval, mean, na.rm=TRUE)

# change array to dateframe, change name of column
stepsInt <- as.data.frame(stepsInt)
names(stepsInt) <- "steps"

# add column for interval
stepsInt$interval <- row.names(stepsInt)

# create plot of mean steps per interval
ggplot(stepsInt, aes(interval, steps, group=1)) + geom_line() + geom_point()
```

## The 5-minute interval that, on average, contains the maximum number of steps

```{r}

# find interval that has maximum number of steps
maxInt <- stepsInt[max(stepsInt$steps), 2]
maxInt
```

## 6. Code to describe and show a strategy for imputing missing data

```{r}
# calculate the number of records with missing steps value
miss <- sum(is.na(stepsDf$steps))
miss

# create copy of stepsDf to add imputed values of steps
stepsImp <- stepsDf

# step through intervals in stepsImp, if steps is NA then substitute 
# the mean steps for that interval
for(i in seq(0, 2355, 5)) {
  stepsImp[is.na(stepsImp$steps) & stepsImp$interval==i, 1] <-
    mean(stepsImp[stepsImp$interval==i, 1], na.rm = TRUE)
}
```

7. Histogram of the total number of steps taken each day after missing values are imputed

```{r}
# repeat section 2 (chunk 3), substituting stepsImp (dataframe with 
# missing values imputed) for stepsDf
stepsDailyImp <- tapply(stepsImp$steps, stepsImp$date, sum)
stepsDailyImp <- as.data.frame(stepsDailyImp)
names(stepsDailyImp) <- "steps"

qplot(stepsDailyImp$steps, geom="histogram")
```

```{r}
# recalculate mean and median number of steps per day after imputed values are added
meanStepsImp <- mean(stepsDailyImp$steps)
meanStepsImp

medStepsImp <- median(stepsDailyImp$steps)
medSteps

# combine the 2 mean and median files to determine what changes imputing missing values made
stepsSummary <- cbind(meanSteps, meanStepsImp, medSteps, medStepsImp)
stepsSummary
```

There was an increase in both the mean and median number of steps per day after missing values were imputed. This means that the missing values tended to be during less active intervals of the day.

8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
# create dayType column in stepsImp, designate it as weekend or weekday based on the date
stepsImp$dayType <- factor(weekdays(stepsImp$date) %in% c("Saturday", "Sunday"), levels=c(FALSE, TRUE), labels=c("weekday","weekend"))

# determine means grouped by interval and weekday/weekend
dayTypeMeans <- aggregate(stepsImp$steps, list(stepsImp$interval,
                    stepsImp$dayType), mean)

# rename columns
names(dayTypeMeans) <- c("interval", "dayType", "steps")

# create a panel plot of mean steps per interval with separate plots for weekdays and weekends
ggplot(dayTypeMeans, aes(interval,steps, group=1)) + geom_point() + geom_line() + facet_wrap(~dayType, nrow=1) + labs(y="Steps", x="Interval")

```
